Verifying a Binary Search Tree with Fine-Grained Locking

### 2021-04-27 concurrency/Verification of Concurrent Binary Search Tree.tex
Give a draft of the introduction part.

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\newcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
\newcommand{\us}{$\mu$s}

@@

\catcode`\@=11
\section{Introduction}
Modern CPU design trends have shifted towards multicore processing over the last two decades. This is because the computing power of computers has reached its limit in single-threaded execution, while the quantity of transistors per single-core CPU are increasing as described by Moore's law \cite{moore}. CPUs have multiple cores now because that is the only way to get more transistors per chip, and achieve a desirable throughput. However, doubling the quantity of cores on one chip does not immediately imply an increase in the performance by a factor of two. Programs that we aim to run in multiple cores have to be carefully designed to efficiently divide the workload into threads, and facilitate non-conflicting access to the shared data by multiple threads in parallel. These challenges is resolved by designing the concurrent data structures, algorithms that provide thread-safe access to the shared data by storing and organizing data in a standard way. Concurrent data structures are now widely employed in real-world applications like database algorithms, file systems, networking, e-commerce, etc. Designing these data structures has been considered as one of the difficult tasks in the software development fields. Despite the careful design, implementation, and thorough unit testing of these algorithms, it is very likely to expect subtle bugs and unexpected behaviors of the program. This is because threads interact with each other in complex and subtle ways such that it is almost impossible to cover every corner case while testing. Therefore, it is important to use a precise way of proving the correctness of concurrent data structures which provides a guarantee that the given algorithm correctly implements the specified concurrent data structures.

@@

\subsection{Ghost State and Global Invariants}
We can prove memory safety and race freedom properties of shared-memory concurrent programs, by tracking the transfer of ownership of shared locations between threads that would happen if we ran the program. But, this idea is insufficient to prove that concurrently running threads are successful to accomplish some task, which is also called functional correctness. For instance, we can prove the memory safety properties of the increment example in  \ref{figure1} through the transfer of ownership of shared variable $x$ between threads. But, we can not guarantee that the value of $x$ will be $2$ after all threads complete their execution, which is demonstrated in \ref{figure1} using separation logic assertion.   
\begin{figure}[htb]
\centering
$\texttt{x = 0;}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\hspace{30px}\pi = \pi_1\ .\ \pi_2$\\
$\begin{array}{l || l}
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{acquire(l);} & \texttt{acquire(l);}\\
\mathit{x \mapsto v \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{x++;} & \texttt{x++;}\\
\mathit{x \mapsto v+1 \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v+1 \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{release(l);} & \texttt{release(l);}\\
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\end{array}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}$
\caption{The increment example annotated with separation logic assertion}
\label{figure1}
\end{figure}
We must preserve the connection between the value of shared resources and the work performed by each thread. A common approach to achieve this is to use \emph{ghost variables}, an auxiliary state introduced later in the proof to track the local information about each thread. They do not appear in the original program. Program in \ref{figure1} can be verified by creating ghost state, which tracks the latest action performed by a thread, for each thread with initial value of $0$, and update with $1$ after each thread increment the value of \texttt{x}. In VST, we can create \emph{ghost state} for any Coq type in the form of an arbitrary \emph{partial commutative monoid} (PCM), a set with a partially defined binary operation that is as associative as it can, commutative, and has a unit, as long as we can describe what happens when two elements of that type are joined together. VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} (\textsf{gname} in Coq) associated with the value $a$, which may be of any type. We will see different kind of ghost states used to verify binary search tree in Section \ref{correctness}.

@@

where invariants are provided in the form of assertion  $\knowInv{i}{P}$, and states that $P$ is maintained as an invariant on the global state with name $i$. The operator $\pvs[][]$ is called ``fancy update" operator,that allows to allocate, open, and close the invariants and the later $\triangleright$ operator is used for impredicativity (i.e. $\knowInv{i}{..\knowInv{i}{P}..}$).
\subsection{Atomic Specifications}
\emph{Linearizability} is one of the most popular traditional correctness condition for concurrent data structures. It is defined in terms of a set of high-level operations, such as insert and remove for a set; if a data structure is linearizable, any operation performed on it tends to happen atomically and in some order that corresponds to the real-time ordering of such operations. For instance, if operation $A$ is completed before operation $B$, then $B$ should logically follow $A$. We can prove the linearizability property of concurrent data structures using \emph{global invariants}. Suppose we have an invariant that said ``there exists a linear history of operations performed on the data structures, stored in some ghost state, and the current  state of the data structure is the one that would result from applying each of those operations in order". Then this directly implies that data structure is linearizable. We can allow a data structure operation to access the content of global invariants, treating that operation as an atomic operation if we are able prove that that operation \emph{appears to execute atomically} at some linearization point. This idea was introduced in the TaDA logic \cite{tada} in the form of \emph{atomic Hoare triples} as shown below, and called as  \emph{Logical atomicity}.    
$$\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$

@@

\section{Related Work}
Our work is based on the concurrent separation logic of VST in the style of Iris \cite{higherorderghoststate} and the logical atomicity from TaDA logic  \cite{tada}. The notion of the \emph{lower} and \emph{upper} bound is taken from the flow interface paper \cite{krishna2017flow}. Our main technical contributions are the verification of two different implementation of binary search tree (using fine-grained locking and lock-free technique) with respect to the same abstract specification, and first c-level mechanized verification of concurrent search-based data structure (i.e. BST). Most of the other concurrent separation logic (CSL)-based verification works are on toy languages, while VST lets us use the same logic on real C programs.

@@

Krishna et al. \cite{krishna2017flow} have presented the proof technique for concurrent search structure templates based on the flow framework and the Iris separation logic, and verified the implementation of concurrent B-tree, hash tables, and linked lists based on the templates. Their work is closely related to our work; we took some inspiration from them in building our ghost state. But, their proofs are on a toy language rather than real C code.
\section{Conclusion}

>>>>

\newcommand\dboxed[1]{\dbox{\ensuremath{#1}}}
\newcommand{\master}[2]{\ensuremath{\mathrm{Master}_{#1}(#2)}}
\newcommand{\snap}[1]{\ensuremath{\mathrm{Snapshot}(#1)}}
\newcommand{\ghost}[2]{\ensuremath{\dboxed{#1}^{#2}}}
\renewcommand{\ifthenelse}[3]{\text{if }#1\text{ then }#2\text{ else }#3}
\newcommand{\us}{$\mu$s}

@@

\catcode`\@=11
\section{Introduction}
\input{intro.tex}
Modern CPU design trends have shifted towards multicore processing over the last two decades. This is because the computing power of computers has reached its limit in single-threaded execution, while the quantity of transistors per single-core CPU are increasing as described by Moore's law \cite{moore}. CPUs have multiple cores now because that is the only way to get more transistors per chip, and achieve a desirable throughput. However, doubling the quantity of cores on one chip does not immediately imply an increase in the performance by a factor of two. Programs that we aim to run in multiple cores have to be carefully designed to efficiently divide the workload into threads, and facilitate non-conflicting access to the shared data by multiple threads in parallel. These challenges is resolved by designing the concurrent data structures, algorithms that provide thread-safe access to the shared data by storing and organizing data in a standard way. Concurrent data structures are now widely employed in real-world applications like database algorithms, file systems, networking, e-commerce, etc. Designing these data structures has been considered as one of the difficult tasks in the software development fields. Despite the careful design, implementation, and thorough unit testing of these algorithms, it is very likely to expect subtle bugs and unexpected behaviors of the program. This is because threads interact with each other in complex and subtle ways such that it is almost impossible to cover every corner case while testing. Therefore, it is important to use a precise way of proving the correctness of concurrent data structures which provides a guarantee that the given algorithm correctly implements the specified concurrent data structures.

@@

\subsection{Ghost State and Global Invariants}
We can prove memory safety and race freedom properties of shared-memory concurrent programs, by tracking the transfer of ownership of shared locations between threads that would happen if we ran the program. But, this idea is insufficient to prove that concurrently running threads are successful to accomplish some task, which is also called functional correctness. For instance, we can prove the memory safety properties of the increment example in  \ref{figure1} through the transfer of ownership of shared variable $x$ between threads. But, we can not guarantee that the value of $x$ will be $2$ after all threads complete their execution, which is demonstrated in \ref{figure1} using separation logic assertion.   
\begin{figure}[htb]
\centering
$\texttt{x = 0;}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\qquad\pi = \pi_1\ .\ \pi_2$\\
$\begin{array}{l || l}
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{acquire(l);} & \texttt{acquire(l);}\\
\mathit{x \mapsto v \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{x++;} & \texttt{x++;}\\
\mathit{x \mapsto v+1 \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v+1 \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{release(l);} & \texttt{release(l);}\\
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\end{array}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}$
\caption{The increment example annotated with separation logic assertion}
\label{figure1}
\end{figure}
We must preserve the connection between the value of shared resources and the work performed by each thread. A common approach to achieve this is to use \emph{ghost variables}, an auxiliary state introduced later in the proof to track the local information about each thread. They do not appear in the original program. Program in \ref{figure1} can be verified by creating ghost state, which tracks the latest action performed by a thread, for each thread with initial value of $0$, and update with $1$ after each thread increment the value of \texttt{x}. In VST, we can create \emph{ghost state} for any Coq type in the form of an arbitrary \emph{partial commutative monoid} (PCM), a set with a partially defined binary operation that is as associative as it can, commutative, and has a unit, as long as we can describe what happens when two elements of that type are joined together. VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} (\textsf{gname} in Coq) associated with the value $a$, which may be of any type. We will see different kind of ghost states used to verify binary search tree in Section \ref{correctness}.

@@

where invariants are provided in the form of assertion  $\knowInv{i}{P}$, and states that $P$ is maintained as an invariant on the global state with name $i$. The operator%%  $\pvs[][]$
is called ``fancy update" operator,that allows to allocate, open, and close the invariants and the later $\triangleright$ operator is used for impredicativity (i.e. $\knowInv{i}{..\knowInv{i}{P}..}$).
\subsection{Atomic Specifications}
\emph{Linearizability} is one of the most popular traditional correctness condition for concurrent data structures. It is defined in terms of a set of high-level operations, such as insert and remove for a set; if a data structure is linearizable, any operation performed on it tends to happen atomically and in some order that corresponds to the real-time ordering of such operations. For instance, if operation $A$ is completed before operation $B$, then $B$ should logically follow $A$. We can prove the linearizability property of concurrent data structures using \emph{global invariants}. Suppose we have an invariant that said ``there exists a linear history of operations performed on the data structures, stored in some ghost state, and the current  state of the data structure is the one that would result from applying each of those operations in order". Then this directly implies that data structure is linearizable. We can allow a data structure operation to access the content of global invariants, treating that operation as an atomic operation if we are able prove that that operation \emph{appears to execute atomically} at some linearization point. This idea was introduced in the TaDA logic \cite{tada} in the form of \emph{atomic Hoare triples} as shown below, and called as  \emph{Logical atomicity}.    
$$\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$$

@@

\section{Related Work}
\label{related}
Our work is based on the concurrent separation logic of VST in the style of Iris \cite{higherorderghoststate} and the logical atomicity from TaDA logic  \cite{tada}. The notion of the \emph{lower} and \emph{upper} bound is taken from the flow interface paper \cite{krishna2017flow}. Our main technical contributions are the verification of two different implementation of binary search tree (using fine-grained locking and lock-free technique) with respect to the same abstract specification, and first c-level mechanized verification of concurrent search-based data structure (i.e. BST). Most of the other concurrent separation logic (CSL)-based verification works are on toy languages, while VST lets us use the same logic on real C programs.

@@

Krishna et al. \cite{krishna2017flow} have presented the proof technique for concurrent search structure templates based on the flow framework and the Iris separation logic, and verified the implementation of concurrent B-tree, hash tables, and linked lists based on the templates. Their work is closely related to our work; we took some inspiration from them in building our ghost state. But, their proofs are on a toy language rather than real C code.
\section{Conclusion}
\label{conclusion}
